The Inconsistency in G\"odel's Ontological Argument: \\ A Success Story for AI in Metaphysics

### 2016-01-28 Papers/InProgress/IJCAI/submission.tex
further improvements

\begin{abstract}
  This paper discusses the discovery of the inconsistency in G\"odel's
  ontological argument as a success story for artificial
  intelligence. Despite the popularity of the argument since the
  appearance of G\"odel's manuscript in 1970, the inconsistency of the
  axioms used in the argument remained unnoticed until 2013, when it
  was detected automatically by the higher-order theorem prover
  \textsc{Leo-II}. Understanding and verifying the refutation
  generated by the prover turned out to be a time-consuming task, and
  its completion, as reported here, required the reconstruction of the
  refutation in the Isabelle proof assistant. The development of an
  improved syntactical hiding for the embedding technique, utilizing
  Isabelle's binding notation mechanism, allows the refutation to be
  presented in a human-friendly way, suitable for non-experts in the
  technicalities of higher-order theorem proving. This brings us a
  step closer to wider adoption of logic-based artificial intelligence
  tools by philosophers.

  The paper reports on further significant improvements regarding full
  automation of the ontological argument in the context of a higher-order modal logic S5 with a
  universal accessibility relation.
\end{abstract}

\section{Introduction}\label{sec:introduction}
Without exaggeration Kurt G\"{o}del's ontological
argument for the existence of God \cite{GoedelNotes,ScottNotes} is
amongst the most discussed formal proofs in modern literature. A rich
body of publications -- including very recent ones -- present,
discuss, assess, criticise, modify and improve G\"{o}del's original
work (see e.g.~Sobel~\shortcite{sobel2004logic} and Oppy~\shortcite{sep-ontological-arguments} and the
references therein).  In philosophy lectures at universities the
argument is regularly presented as a masterpiece argument in
metaphysics. Since 2013, when Benzm\"uller and Woltzenlogel-Paleo~\shortcite{J30,C40} first
reported their successful initial computer-assisted
analysis of G\"odel's proof and Scott's variant,
their work has received a media repercussion on a global scale\footnote{A
  small collection of news articles is available at {\scriptsize
    \url{https://github.com/FormalTheology/GoedelGod/blob/master/Press/LinksToNews.md}}},
and numerous bloggers commented on the proof
\cite{fuhrmann15:_blogg_goedel}.

@@

On the technical side, the quest for constructing a compelling
refutation, capable of convincing also human non-experts, led to an
improvement of the syntax of the embedding of modal logics in Isabelle/HOL
(as discussed in \S\ref{sec:improvedsyntax}). With the new syntax, a
(nearly) perfect match between the original pen and paper
presentations and our encoding in Isabelle/HOL is feasible. A more
user-friendly syntax, as reported here, is clearly an important
prerequisite for promoting the theorem proving technology employed
here to a wider community of philosophers, who are not necessarily
experts in automated reasoning or higher-order logic.

Another novel contribution reported here (in \S\ref{sec:improvedembedding}) is
the implementation of an alternative embedding for the slightly
stronger modal logic \SFiveU. Our experiments have shown that the new
embedding is more efficient, as the following two, previously open problems can now be solved:
\begin{itemize}
\item Proving the final theorem T3 \textit{``Necessarily, there
    exists God''} (in Scott's \shortcite{ScottNotes} consistent variant) fully automatically and directly from the axioms
  alone, without relying on the argument's intermediate argumentation
  steps (i.e., lemmata). \marginpar{ToDo: 2004 is the year of Sobel's book, not of Scott's manuscript. We need to fix this as I fixed the date for Goedel's manuscript. But I don't know the correct date for Scott's manuscript.}
\item Verifying automatically (in Isabelle/HOL) the proof of the modal
  collapse \cite{Sobel}, which is one of the most strongly criticized
  `side-effects' of the argument's axioms.
\end{itemize}

@@

St. Anselm's ontological argument \cite{Proslogion} can be regarded as
the ancestor of modern ontological arguments such as G\"odel's. In the
millenium between Anselm and G\"odel, many philosophers modified and
arguably improved Anselm's argument. Of particular importance to
G\"odel was the work of Leibniz. 
%\cite{ToDo:Leibniz}.  
G\"odel's manuscript
(Fig.~\ref{GoedelScript}(a)) can be considered a translation of Leibniz's
presentation of the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_S5U}. The main difference to G\"odel's version is an extra conjunct in the definition of \emph{essence}. G\"odel's
different definition of essence can be seen either in his manuscript
(Fig.~\ref{GoedelScript}(a)) or, in more modern notation, in the Isabelle
formalization shown in Fig.~\ref{InconsistencyIsabelleK}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required. This difference has been
considered inessential and merely an oversight by many. For instance,
Hazen [\citeyear[p.365]{Hazen}] states that ``G\"odel left this
clause out in [11], but this appears to have been an oversight''.
% -- it is
%included in related manuscripts''. 
However, the extra conjunct is in
fact crucial. Without it, G\"odel's original axioms are
inconsistent. With it, Scott's axioms are consistent (cf.~Fig.~\ref{Scott_S5U}
where the model finder Nitpick \cite{Nitpick} confirms consistency).\footnote{In
  personal communication, Dana Scott confirmed that he was unaware
  that G\"odel's axioms were inconsistent.}

@@

For more than four decades, the serious consequences of G\"odel's
mistake remained unnoticed, despite intense activity focused on
criticizing his argument. Especially since the discovery by Sobel
\shortcite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
  $\phi\rightarrow \Box \phi$, states that contingent truth implies
  necessary truth; it can be interpreted as \textit{everything is
    pre-determined} or even \textit{there is no free will}. } is
entailed by G\"odel's (or also Scott's) axioms, several variants have
been proposed
\cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
attempting to avoid the modal collapse. Many of these variants omit
the crucial conjunct in the definition of essence as well\footnote{As
  these variants also change other axioms, on which the inconsistency
  of G\"odel's axioms depends, it is not necessarily the case that
  these variants are also inconsistent; they to be analyzed
  separately.}. Opponents of the argument
(e.g. Oppy [\citeyear[p.226--227]{oppy96:_goedel_ontol_argum};
\citeyear[p.364]{oppy00:_respon_gettin};
\citeyear[p.1068]{oppy08:_higher_order_ontol_argum}]) have also proposed
parodies and other criticisms, referring to variants where the
conjunct is omitted.

@@

For a computational analysis of G\"odel's ontological argument, the embedding of HOMLs such as K,
KB and S5 with various domain conditions (possibilist and actualist quantification)
is required. This idea has been successfully employed in related work
\cite{C40}. The embedding of HOML is in fact
straightforward. Formulas in HOML are \emph{lifted}, translated to predicates
over worlds, which are themselves explicitly represented as
terms. The logical constants of HOML are translated to HOL terms in such a way that, for instance, 
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_o$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation
\cite{DBLP:journals/logcom/Ohlbach91}, which is here intra-logically
realized --- and extended for quantifiers --- in HOL by stating a set
of equations defining the logical constants (Fig.~\ref{QML_S5U}). The
resulting object logic is the HOML K with rigid terms and constant
domains (possibilist quantifiers). Other logics (e.g. KB, S5) can be
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by
means of an existence predicate, which can be used to guard the
quantifiers. The embedding approach is, therefore, very flexible.

\subsection{Improved Embedding}\label{sec:improvedembedding}
\begin{figure}[t]
\centerline{\includegraphics[width=\columnwidth]{./Images/QML_S5U.png}}
\caption{Improved Embedding of \SFiveU} \label{QML_S5U}
\end{figure}

The modal logic \SFive requires that the accessibility relation be
reflexive, symmetric and transitive. The usual approach to embed
\SFive would be to use the standard translation for K described above
and to state that $r$ is an equivalence relation, e.g., by postulating
the following axioms:
\begin{itemize}
\item Reflexivity: $\forall x. (r~x~x)$
\item Symmetry: $\forall x. \forall y. (r~x~y) \rightarrow (r~y~x)$ 
\item Transitivity: $\forall x. \forall y. \forall z. (r~x~y) \wedge (r~y~z) \rightarrow (r~x~z)$
\end{itemize}
Instead, we consider here a modal logic that we call \SFiveU, which is
characterized by the following condition on $r$:
\begin{itemize}
\item Universality: $\forall x. \forall y. (r~x~y)$
\end{itemize}

@@

The new embedding of \SFiveU in Isabelle is shown in
Fig.~\ref{QML_S5U}. It is important to note that we 
have  $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi$,
despite the differences in the underlying frame structures.

%
% \marginpar{@Bruno: I commented you paragraph here, since I am not
%   convinced (yet) that e.g. BF and CBF are equally valid in \SFiveU
%   and \SFive}
% \SFiveU is only slightly stronger than \SFive. Most importantly,
% $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi^{U}$, where
% $\varphi^U$ is obtained from $\varphi$ by replacing all $\Box$ and
% $\Dia$ by, respectively, $\BlackBox$ and $\BlackDia$. Therefore,
% \SFiveU can be considered as adequate for metaphysics as \SFive
% (cf. \cite[p. 127]{WilliamsonModalLogicAsMetaphysics}). 
% % \cite{Mattey}, \cite[ToDo]{DunnHardegree}
%
With this improved embedding, the final theorem T3
(\textit{``Necessarily, there exists God''}) can be derived from
Scott's consistent version of the axioms fully automatically.  The fully
automatic proof has been generated (in about 2,5 seconds) by the theorem prover
\textsc{Leo-II}~\cite{C26} and subsequently verified in the proof
assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{Scott_S5U}.
The collaboration between the two systems has been orchestrated by
Isabelle's Sledgehammer tool \cite{Sledgehammer}.

@@

\begin{figure}[t]
\centerline{\includegraphics[width=\columnwidth]{./Images/Scott_S5U.png}}
\caption{Full Automation of T3 in \SFiveU; Consistency of Scott's
  Axioms;  Automatic Verification of Modal Collapse} \label{Scott_S5U}
\end{figure}

@@

Wider adoption of higher-order theorem proving technology for reasoning about and within embedded object logics, especially among non-expert users, is still hindered by the gap between the syntax used by people, when they write logical formulas with pen and paper, and the syntax used by higher-order logic theorem provers. Even when the syntax of the underlying higher-order system is elegant (as is the case in Isabelle/HOL), the embedding of HOML into HOL may easily expose details of HOL that may be uncommon to the user, disturbing his/her experience while using the system. To illustrate this point, Fig.~\ref{UglyEssence} shows how the definition of essence looked like in previous work \cite{J28}, where advanced syntax-sugaring features were not used. It looks notably higher-order, and its style differs significantly from the common style seen in works on modal logics and the ontological argument. The following specific issues can be enumerated:
\begin{enumerate}
\item Lambda abstractions (which are a typical higher-order feature) appear explicitly in places where they didn't need to appear in a pure HOML formulation (cf. G\"odel's manuscript, Fig.~\ref{GoedelScript}(a)).
\item Quantifiers appear as higher-order defined constants, and not as binders. This forces the user to read (and write) formulas of the form $\forall (\lambda x. A(x))$ instead of the more common $\forall x. A(x)$.
\item The lifted modal connectives are represented by prefixing the
  letter ``m'' (e.g. $m\wedge$ and $m\imp$). The prefix disturbs the
  user, as it constantly reminds him/her that there is something
  unusual about the modal connectives.
\item Higher-order parenthesis conventions for the application of a predicate to a term are used. 
Instead of reading $\psi(y)$, as he would expect, he has to read $(\psi ~ y)$. Outside niche areas in computer science, the former syntax is more widely known than the latter.
\end{enumerate}

\begin{figure}
\centerline{\includegraphics[width=\columnwidth]{./Images/UglyEssence.png}}
\caption{ Definition of Essence using Old Syntax } \label{UglyEssence}
\end{figure}

@@

\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{./Images/LEO-Proof.png}}
\caption{Lines 115--120 of \textsc{Leo-II}'s refutation. Primitive
  substitutions (e.g. with the empty property) are highlighted.} \label{LEO-Proof}
\end{figure*}

Although \textsc{Leo-II}'s resolution refutation is not easy to read
for humans, it did contain relevant hints to the importance of the
empty property $\lambda x. \bot$ (cf. Fig.~\ref{LEO-Proof}).
%
Note that the terms for the empty property ($\lambda x. \bot$) and for the property of self-difference ($\lambda x.  x\not=x$) have identical denotations in a logic setting
with full functional and Boolean extensionality as given
here. Nevertheless, some philosophers may actually prefer the use of
self-difference over the empty property\footnote{This occurred in
  a private communication of the authors with Andr\'e Fuhrmann.} in
the analysis below. However, for the proof to go through it is
irrelevant which notion we use and the reader may simply replace the
empty property by self-difference. In HOL it is customary to think of a set as a unary predicate that is true for (and only for) elements of the set. Therefore, the empty property (i.e. the unary predicate that is false for any individual) represents the empty set, and hence we use the empty set symbol ($\emptyset$) as an abbreviation for the empty property. \marginpar{ToDo: should we use the empty set symbol in the isabelle code in figure 5 as well? Or should we instead stop using the empty set symbol in the informal argument (and use $\lambda x. \bot$ instead) and remove this convoluted commentary that I added here.}

@@

\begin{enumerate}
\item From G\"odel's definition of essence 
(${\ess{\phi}{x} \biimp {\allq \psi} (\psi(x)
\imp {\nec} \allq y (\phi(y) \imp \psi(y)))}$) it follows that the
empty property (or self-difference) is an essence of every individual (\textbf{Empty Essence Lemma}): 
$$\allq x\,(\ess{\emptyset}{x})$$

\item From theorem T1 (positive Properties are possibly
  exemplified: ${\allq \phi} [P(\phi) \imp {\pos}  \exq x
  \phi(x)]$) and axiom A5 (``necessary existence'' is a positive property: $P(\NE)$ ), it follows that $\NE$ is possibly exemplified:
  $$
  \pos \exq x [\NE(x)]
  $$

@@

Interestingly, the refutation automatically generated by
\textsc{Leo-II} uses a symmetric accessibility relation, and thus
requires the modal logic KB. The informal, human-constructed
refutation described above, on the other hand, requires only the
weaker modal logic K. In our experiments \textsc{Leo-II} (like all
other HOL provers) was still too weak to automatically prove the
inconsistency already in logic K. Hence, this remains an open problem for automated
theorem provers.

\subsection{Argument Reconstruction in Isabelle}  \label{sec:arg2}
To verify the correctness of the informal argument explained above, it
was reconstructed in Isabelle/HOL, using Metis\footnote{Metis, unlike
  external provers such as \textsc{Leo-II} or Satallax, 
  constructs proofs in Isabelle's highly trusted kernel calculus.} to automate the
inessential parts. The essential use of the Empty Essence Lemma, on
the other hand, is explicitly stated, to ensure that Isabelle is
reconstructing the same argument. In fact, without the help of this
lemma, Metis is still not strong enough to refute G\"odel's
axioms.
\begin{figure}[t]
\centerline{\includegraphics[width=\columnwidth]{./Images/InconsistencyIsabelleK.png}}
\caption{Refutation of G\"odel's Axioms in Isabelle/HOL.} \label{InconsistencyIsabelleK}
\end{figure}

@@

Isabelle/HOL's Sledgehammer tool, which orchestrates the calls to
external automated theorem provers, such as \textsc{Leo-II}, still
fails to detect the inconsistency of G\"odel's axioms in the standard
embedding, while a direct modeling of the problem in TPTP THF syntax
in combination with a direct call of \textsc{Leo-II} succeeds. In
other words, without independent experiments directly in TPTP THF the
inconsistency would thus not have been detected.
\begin{figure*}
\centerline{
\includegraphics[width=.65\textwidth]{./Images/Manuscript2.png} \hfill
\includegraphics[width=.35\textwidth]{./Images/Inconsistency_S5U_direct.png}
}
\caption{(a) G\"{o}del's manuscript, with mutually inconsistent axioms and
  definitions highlighted; (b) Refutation in
\SFiveU} \label{Inconsistency_S5U} \label{GoedelScript} 
\end{figure*}

\marginpar{ToDo: Are we really supposed to use \$ instead of ``Sec.''?}
On the other hand (and further confirming the claims from \$\ref{sec:improvedembedding}), 
the reconstruction in Isabelle/HOL with the improved embedding for \SFiveU was more efficient: 
the inconsistency could be detected
by \textsc{Leo-II} also when called via
Sledgehammer. Moreover, the result could subsequently be verified with
Metis even without the Empty Essence Lemma; cf. Fig.\ref{Inconsistency_S5U}. 
Interestingly, Metis' argument is different from that of
\$\ref{sec:arg1} and \$\ref{sec:arg2}, since axiom A1b is not used.

\section{Conclusion}\label{sec:conclusion}
The axioms and definitions in G\"odel's manuscript are inconsistent;
this has been detected automatically by the prover
\textsc{Leo-II}. Here we have presented a rational reconstruction and
verification of the inconsistency argument in Isabelle/HOL. This
argument is valid in all normal HOMLs including base logic K.

>>>>

\begin{abstract}
  This paper discusses the discovery of the inconsistency in G\"odel's
  ontological argument as a success story for artificial
  intelligence. Despite the popularity of the argument since the
  appearance of G\"odel's manuscript in 1970, the inconsistency of the
  axioms used in the argument remained unnoticed until 2013, when it
  was detected automatically by the higher-order theorem prover
  \textsc{Leo-II}. Understanding and verifying the refutation
  generated by the prover turned out to be a time-consuming task, and
  its completion, as reported here, required the reconstruction of the
  refutation in the Isabelle proof assistant. The development of an
  improved syntactical hiding for the embedding technique
  % , utilizing
  % Isabelle's binding notation mechanism, 
  allows the refutation to be
  presented in a human-friendly way, suitable for non-experts in the
  technicalities of higher-order theorem proving. This brings us a
  step closer to wider adoption of logic-based artificial intelligence
  tools by philosophers.

  In addition, the paper reports on further improvements regarding the
  full automation of the ontological argument in a higher-order modal
  logic \textbf{S5} with a universal accessibility relation.
\end{abstract}

\section{Introduction}\label{sec:introduction}
Without exaggeration Kurt G\"{o}del's ontological
argument for the existence of God \cite{GoedelNotes,ScottNotes} is
amongst the most discussed formal proofs in modern literature. A rich
body of publications -- including very recent ones -- present,
discuss, assess, criticise, modify and improve G\"{o}del's original
work (see e.g.~Sobel~\shortcite{sobel2004logic} and Oppy~\shortcite{sep-ontological-arguments} and the
references therein).  In philosophy lectures at universities the
argument is regularly presented as a masterpiece argument in
metaphysics. Since 2013, when Benzm\"uller and Woltzenlogel-Paleo~\shortcite{J30,C40} first
reported their successful initial computer-assisted
analysis of G\"odel's proof and Scott's variant,
their work has received a media repercussion on a global scale\footnote{A
  collection of news articles is available at {\scriptsize
    \url{https://github.com/FormalTheology/GoedelGod/blob/master/Press/LinksToNews.md}}},
and numerous bloggers commented on the proof
\cite{fuhrmann15:_blogg_goedel}.

@@

On the technical side, the quest for constructing a compelling
refutation, capable of convincing also human non-experts, led to an
improvement of the syntax of the embedding of modal logics in Isabelle/HOL
(as discussed in \S\ref{sec:improvedsyntax}). With the new syntax, a
(nearly) perfect match between the original pen and paper
presentations and our encoding in Isabelle/HOL is feasible. A more
user-friendly syntax, as reported here, is clearly an important
prerequisite for promoting the theorem proving technology employed
here to a wider community of philosophers, who are not necessarily
experts in automated reasoning or HOL.

Another novel contribution reported here (in
\S\ref{sec:improvedembedding}) is the implementation of an alternative
embedding for the slightly stronger modal logic \SFiveU, which is
based on a universal accessibility relation. Our experiments have
shown that the new embedding is more efficient, as the following two,
previously open problems can now be solved:
\begin{itemize}
\item Proving automatically the final theorem T3 \textit{Necessarily, there
    exists God} (in Scott's \shortcite{ScottNotes} consistent variant) directly from the axioms
  alone, without relying on the argument's intermediate argumentation
  steps (i.e., lemmata). \marginpar{ToDo: 2004 is the year of Sobel's book, not of Scott's manuscript. We need to fix this as I fixed the date for Goedel's manuscript. But I don't know the correct date for Scott's manuscript.}
\item Verifying automatically (in Isabelle/HOL) the proof of the modal
  collapse \cite{Sobel}, which is one of the most strongly criticized
  `side-effects' of the argument's axioms.
\end{itemize}

@@

St. Anselm's ontological argument \cite{Proslogion} can be regarded as
the ancestor of modern ontological arguments such as G\"odel's. In the
millenium between Anselm and G\"odel, many philosophers modified and
arguably improved Anselm's argument. Of particular importance to
G\"odel was the work of Leibniz. 
%\cite{ToDo:Leibniz}.  
G\"odel's manuscript
(Fig.~\ref{GoedelScript}(a)) can be considered a translation of Leibniz's
presentation of the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_S5U}. The main difference to G\"odel's version is an extra conjunct in the definition of \emph{essence}. G\"odel's
different definition of essence can be seen either in his manuscript
(Fig.~\ref{GoedelScript}(a)) or, in more modern notation, in the Isabelle
formalization shown in Fig.~\ref{InconsistencyIsabelleK}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required. This difference has been
considered inessential and merely an oversight by many. For instance,
Hazen [\citeyear[p.365]{Hazen}] states that ``G\"odel left this
clause out in [11], but this appears to have been an oversight''.
% -- it is
%included in related manuscripts''. 
However, the extra conjunct is in
fact crucial. Without it, G\"odel's original axioms are
inconsistent. With it, Scott's axioms are consistent (cf.~Fig.~\ref{Scott_S5U}
where the model finder Nitpick \cite{Nitpick} confirms consistency).\footnote{In
  personal communication, Dana Scott confirmed that he was unaware
  at the time that G\"odel's axioms were inconsistent.}

\begin{figure}[t]
\centerline{\includegraphics[width=.95\columnwidth]{./Images/Scott_S5U.png}}
\caption{Full Automation of T3 in \SFiveU; Consistency of Scott's
  Axioms;  Automatic Verification of Modal Collapse} \label{Scott_S5U}
\end{figure}

@@

For more than four decades, the serious consequences of G\"odel's
mistake remained unnoticed, despite intense activity on
criticizing his argument. Especially since the discovery by Sobel
\shortcite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
  $\phi\rightarrow \Box \phi$, states that contingent truth implies
  necessary truth; it can be interpreted as \textit{everything is
    pre-determined} or even \textit{there is no free will}. } is
entailed by G\"odel's (or also Scott's) axioms, several variants have
been proposed
\cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
attempting to avoid the modal collapse. Many of these variants omit
the crucial conjunct in the definition of essence as well.\footnote{As
  these variants also change other axioms, on which the inconsistency
  of G\"odel's axioms depends, it is not necessarily the case that
  these variants are also inconsistent; they to be analyzed
  separately.} Opponents of the argument
(e.g. Oppy [\citeyear[p.226--227]{oppy96:_goedel_ontol_argum};
\citeyear[p.364]{oppy00:_respon_gettin};
\citeyear[p.1068]{oppy08:_higher_order_ontol_argum}]) have also proposed
parodies and other criticisms, referring to variants where the
conjunct is omitted.

@@

For a computational analysis of G\"odel's ontological argument, the embedding of HOMLs such as \textbf{K},
\textbf{KB} and \textbf{S5} with various domain conditions (possibilist and actualist quantification)
is required. This idea has been successfully employed in related work
\cite{C40}. The embedding of HOML is in fact
straightforward. Formulas in HOML are \emph{lifted}, i.e., converted into predicates
over worlds, which are themselves explicitly represented as
terms. The logical constants of HOML are translated to HOL terms in such a way that, for instance, 
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_o$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation
\cite{DBLP:journals/logcom/Ohlbach91}, which is here intra-logically
realized --- and extended for quantifiers --- in HOL by stating a set
of equations defining the logical constants (Fig.~\ref{QML_S5U}). The
resulting object logic is the HOML \textbf{K} with rigid terms and constant
domains (possibilist quantifiers). Other logics (e.g. \textbf{KB}, \textbf{S5}) can be
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by
means of an existence predicate, which can be used to guard the
quantifiers. The embedding approach is, therefore, very flexible.

\subsection{Improved Embedding}\label{sec:improvedembedding}
\begin{figure}[t]
\centerline{\includegraphics[width=.95\columnwidth]{./Images/QML_S5U.png}}
\caption{Improved Embedding of \SFiveU} \label{QML_S5U}
\end{figure}

The modal logic \SFive requires that the accessibility relation be
reflexive, symmetric and transitive. The usual approach to embed
\SFive would be to use the standard translation for \textbf{K} described above
and to state that $r$ is an equivalence relation, e.g., by postulating
the following axioms:
\begin{itemize}
\item Reflexivity: $\forall x. (r~x~x)$
\item Symmetry: $\forall x. \forall y. (r~x~y) \rightarrow (r~y~x)$ 
\item Transitivity: $\forall x. \forall y. \forall z. (r~x~y) \wedge (r~y~z) \rightarrow (r~x~z)$
\end{itemize}
Instead, we consider here a modal logic that we call \SFiveU, which is
characterized by the following condition on $r$:
\begin{itemize}
\item Universality: $\forall x. \forall y. (r~x~y)$
\end{itemize}

@@

The new embedding of \SFiveU in Isabelle/HOL is shown in
Fig.~\ref{QML_S5U}. It is important to note that we 
have  $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi$,
despite the differences in the underlying frame structures.

%
% \marginpar{@Bruno: I commented you paragraph here, since I am not
%   convinced (yet) that e.g. BF and CBF are equally valid in \SFiveU
%   and \SFive}
% \SFiveU is only slightly stronger than \SFive. Most importantly,
% $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi^{U}$, where
% $\varphi^U$ is obtained from $\varphi$ by replacing all $\Box$ and
% $\Dia$ by, respectively, $\BlackBox$ and $\BlackDia$. Therefore,
% \SFiveU can be considered as adequate for metaphysics as \SFive
% (cf. \cite[p. 127]{WilliamsonModalLogicAsMetaphysics}). 
% % \cite{Mattey}, \cite[ToDo]{DunnHardegree}
%
With this improved embedding, the final theorem T3
(\textit{Necessarily, there exists God}) can be derived from
Scott's consistent version of the axioms fully automatically.  The fully
automatic proof has been generated (in about 2,5 seconds) by the theorem prover
\textsc{Leo-II}~\cite{C26} and subsequently verified in the proof
assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{Scott_S5U}.
The collaboration between the two systems has been orchestrated by
Isabelle's Sledgehammer tool \cite{Sledgehammer}.

@@

@@

Wider adoption of HOL theorem proving technology for
reasoning about and within embedded object logics, especially among
non-expert users, is still hindered by the gap between the syntax used
by people, when they write logical formulas with pen and paper, and
the syntax used by HOL theorem provers. Even when the syntax of the
underlying higher-order system is elegant (as is the case in
Isabelle/HOL), the embedding of HOML into HOL may easily expose
details of HOL that may be uncommon to the user, disturbing his/her
experience while using the system. To illustrate this point,
Fig.~\ref{UglyEssence} shows how the definition of essence looked like
in previous work \cite{J28}, where advanced syntax-sugaring features
were not used. It looks notably higher-order, and its style differs
significantly from the common style seen in works on modal logics and
the ontological argument. The following specific issues can be
enumerated:
\begin{enumerate}
\item $\lambda$-abstractions, a typical higher-order feature, appear explicitly in places where they didn't need to in a pure HOML formulation (cf. G\"odel's manuscript, Fig.~\ref{GoedelScript}(a)).
\item Quantifiers appear as higher-order defined constants, and not as binders. This forces the user to read (and write) formulas of the form $\forall (\lambda x. A(x))$ instead of the more common $\forall x. A(x)$.
\item The lifted modal connectives are represented by prefixing the
  letter ``m'' (e.g. $m\wedge$ and $m\imp$). The prefix disturbs the
  user, as it constantly reminds him/her that there is something
  unusual about the modal connectives.
\item Higher-order parenthesis conventions for the application of a predicate to a term are used. 
Instead of reading $\psi(y)$, as he would expect, he has to read $(\psi ~ y)$. Outside niche areas in computer science, the former syntax is more widely known than the latter.
\end{enumerate}

\begin{figure}
\centerline{\includegraphics[width=.95\columnwidth]{./Images/UglyEssence.png}}
\caption{ Definition of Essence using Old Syntax } \label{UglyEssence}
\end{figure}

@@

\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{./Images/LEO-Proof.png}}
\caption{Lines 115--120 of \textsc{Leo-II}'s refutation. Primitive
  substitutions (e.g. with the empty property) are highlighted. In the
red part, property variable $\texttt{SV8}$ has been instantiated with the
$\lambda \texttt{SV16}. \lambda \texttt{SV17}. \bot$, i.e., the empty property.
} \label{LEO-Proof}
\end{figure*}

Although \textsc{Leo-II}'s resolution refutation is not easy to read
for humans, it did contain relevant hints to the importance of the
empty property $\lambda x. \bot$ (respectively $\lambda x. \lambda
y. \bot$ in the embedding approach, cf. Fig.~\ref{LEO-Proof}; the 
additional ``$\lambda y.$'' is due to the formula lifting).
%
Note that the terms for the empty property ($\lambda x. \bot$) and for the property of self-difference ($\lambda x.  x\not=x$) have identical denotations in a logic setting
with full functional and Boolean extensionality as given
here. Nevertheless, some philosophers may actually prefer the use of
self-difference over the empty property\footnote{This occurred in
  a private communication of the authors with Andr\'e Fuhrmann.} in
the analysis below. However, for the proof to go through it is
irrelevant which notion we use and the reader may simply replace the
empty property by self-difference. In HOL it is customary to think of
a set as a unary predicate that is true for (and only for) elements of
the set. Therefore, the empty property (i.e. the unary predicate that
is false for any individual) represents the empty set, and hence we
use the empty set symbol $\emptyset$ as an abbreviation for the empty property. \marginpar{ToDo: should we use the empty set symbol in the isabelle code in figure 5 as well? Or should we instead stop using the empty set symbol in the informal argument (and use $\lambda x. \bot$ instead) and remove this convoluted commentary that I added here.}

@@

\begin{enumerate}
\item From G\"odel's definition of essence 
(${\ess{\phi}{x} \biimp {\allq \psi} (\psi(x)
\imp {\nec} \allq y (\phi(y) \imp \psi(y)))}$) it follows that the
empty property (or self-difference) is an essence of every individual (\textbf{Empty Essence Lemma}): 
$$\allq x (\ess{\emptyset}{x})$$

\item From theorem T1 (\textit{Positive properties are possibly
  exemplified}: ${\allq \phi} [P(\phi) \imp {\pos}  \exq x
  \phi(x)]$) and axiom A5 (``necessary existence'' is a positive property: $P(\NE)$ ), it follows that $\NE$ is possibly exemplified:
  $$
  \pos \exq x [\NE(x)]
  $$

@@

Interestingly, the refutation automatically generated by
\textsc{Leo-II} uses a symmetric accessibility relation, and thus
requires the modal logic \KB. The informal, human-constructed
refutation described above, on the other hand, requires only the
weaker modal logic \K. In our experiments \textsc{Leo-II} (like all
other HOL provers) was still too weak to automatically prove the
inconsistency already in logic \K. Hence, this remains an open problem for automated
theorem provers.

\subsection{Argument Reconstruction in Isabelle}  \label{sec:arg2}
To verify the correctness of the informal argument explained above, it
was reconstructed in Isabelle/HOL, using Metis\footnote{Metis, unlike
  external provers such as \textsc{Leo-II} or Satallax, 
  constructs proofs in Isabelle's highly trusted kernel calculus.} to automate the
inessential parts. The essential use of the Empty Essence Lemma, on
the other hand, is explicitly stated, to ensure that Isabelle is
reconstructing the same argument. In fact, without the help of this
lemma, Metis is still not strong enough to refute G\"odel's
axioms.
\begin{figure}[t]
\centerline{\includegraphics[width=.95\columnwidth]{./Images/InconsistencyIsabelleK.png}}
\caption{Refutation of G\"odel's Axioms in Isabelle/HOL.} \label{InconsistencyIsabelleK}
\end{figure}

@@

Isabelle/HOL's Sledgehammer tool, which orchestrates calls to
external provers such as \textsc{Leo-II}, still
fails to detect the inconsistency of G\"odel's axioms (in the standard
embedding), while a direct modeling of the problem in TPTP THF syntax
in combination with a direct call of \textsc{Leo-II} succeeded. In
other words, without independent experiments directly in TPTP THF the
inconsistency would thus not have been detected.
\begin{figure*}
\centerline{
\includegraphics[width=.65\textwidth]{./Images/Manuscript2.png} \hfill
\includegraphics[width=.35\textwidth]{./Images/Inconsistency_S5U_direct.png}
}
\caption{(a) G\"{o}del's manuscript, with mutually inconsistent axioms and
  definitions highlighted; (b) Refutation in
\SFiveU} \label{Inconsistency_S5U} \label{GoedelScript} 
\end{figure*}

%\marginpar{ToDo: Are we really supposed to use \S instead of ``Sec.''?}
On the other hand (and further confirming the claims from \S\ref{sec:improvedembedding}), 
the reconstruction in Isabelle/HOL with the improved embedding for \SFiveU was more efficient: 
the inconsistency could be detected
by \textsc{Leo-II} also when called via
Sledgehammer. Moreover, the result could subsequently be verified with
Metis even without the Empty Essence Lemma; cf. Fig.\ref{Inconsistency_S5U}. 
Interestingly, Metis' argument is different from that of
\S\ref{sec:arg1} and \S\ref{sec:arg2}, since axiom A1b is not used.

\section{Conclusion}\label{sec:conclusion}
The axioms and definitions in G\"odel's manuscript are inconsistent;
this has been detected automatically by the prover
\textsc{Leo-II}. Here we have presented a rational reconstruction and
verification of the inconsistency argument in Isabelle/HOL. This
argument is valid in all normal HOMLs including base logic \K.
